Genome-wide changes in expression during inflorescence architecture development

### 2015-09-04 ms.Rmd
typo in hypergeom code

Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`). In general, the GSEA results correspond to the hypergeometric tests for transcription factor enrichment in the clusters. For example, SBP transcription factors are enriched in RM and PBM, consistent with their overrepresentation in cluster `r getClusterName('small', 'dec')`, which contains genes with higher expression in those meristem types.

@@

These patterns are not intended to represent each member of each gene family. For example, there are several MADS genes that have lower expression in the SM than in the RM, but expressed MADS genes were detected at a higher level overall in SM. Despite this caveat, there are some notable modes of expression shared between different families that may indicate roles in inflorescence development. In the GSEA, the most pronounced change in expression appears to occur between the RM and branch meristem samples. Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types. Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.

@@

Analysis of the sequenced libraries was carried out with custom `bash` and `R` scripts [@RALanguageandEn:2015wf], which were arranged into a pipeline using the `ruffus` package for `python3` [@Goodstadt:2010jp]. All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and paramaters of the software used, are available with revision history in a public GitHub repository at https://github.com/evoreprice/lmdPaper.

Briefly, adaptor trimming was performed with `cutadapt` [@Martin:2011eu] and reads were mapped against the MSUv7 genome and annotation [@Kawahara:2013gh; @Ouyang:2007il] downloaded from Phytozome 10.3 [@Goodstein:2012ho] using `STAR` [@Dobin:2012fg] in 2-pass mode. The `'quantMode'` argument of `STAR` was used for read counting. rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped. Differential expression analysis to calculate Log~2~ fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do]. To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf]. These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distrubution to those used for calculating the 'genic' expression values. For each library, the cutoff was placed at the 95^th^ percentile of the distribution of intergenic expression values. Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package [@Kumar:2007uw] and geneset enrichment analysis with the `gage` package [@Luo:2009kp] respectively. All plots were produced with the `ggplot2` package [@Wickham:2009tn].

@@

**`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**. Geneset enrichment analysis (GSEA) for transcription factor and other transcriptional regulator families. Test statistics were calculated on a matrix of Log~2~ fold change values between each sample and the other three samples for expressed family members. A positive test statistic indicates enrichment and a negative statistic indicates depletion. Although some test statistics had low *p*-values none were lower than 0.10 after Benjamini-Hochberg correction.

>>>>

Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`). Unlike soft clustering, which groups genes based on shared expression dynamics, this GSEA was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in the expression of their members in each stage. These patterns are not intended to represent each member of each gene family: for example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM. Despite this caveat, there are some notable modes of expression shared between different families that may indicate roles in inflorescence development. The most pronounced change in expression appears to occur between the RM and branch meristem samples. Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types. Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.

@@

@@

Analysis of the sequenced libraries was carried out with `bash` and `R` scripts [@RALanguageandEn:2015wf], which were arranged into a pipeline using the `ruffus` package for `python3` [@Goodstadt:2010jp]. All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and paramaters of the software used, are available with revision history in a public GitHub repository at https://github.com/evoreprice/lmdPaper.

Briefly, adaptor trimming was performed with `cutadapt` [@Martin:2011eu] and reads were mapped against the MSUv7 genome and annotation [@Kawahara:2013gh; @Ouyang:2007il] downloaded from Phytozome 10.3 [@Goodstein:2012ho] using `STAR` [@Dobin:2012fg] in 2-pass mode. The `'quantMode'` argument of `STAR` was used for read counting. rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped. Differential expression analysis to calculate Log~2~-fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do]. To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf]. These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distrubution to those used for calculating the 'genic' expression values. For each library, the cutoff was placed at the 95^th^ percentile of the distribution of intergenic expression values. Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package [@Kumar:2007uw] and geneset enrichment analysis with the `gage` package [@Luo:2009kp] respectively. All plots were produced with the `ggplot2` package [@Wickham:2009tn].

@@

**`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**. Geneset enrichment analysis (GSEA) for transcription factor and other transcriptional regulator families. Test statistics were calculated from the L~2~FC values between the stage of interest and all other stages for the expressed genes in each geneset. The absolute value of the test statistic indicates the magnitude of the geneset-level change, and a positive test statistic indicates enrichment and a negative statistic indicates depletion. Although some test statistics had low *p*-values none were lower than 0.10 after Benjamini-Hochberg correction.
